Left-Linear Rewriting in Adhesive Categories

### 2024-07-04 concur/final.tex
minor on rules
minima pro formattazione

\usepackage{pgf}
\usepackage{tikz}
\usepackage{tikz-cd}
\usetikzlibrary{arrows,shapes,snakes,automata,backgrounds,petri,fit,positioning}
\tikzstyle{node}=[circle, draw=black, minimum size=1mm, inner sep=1.5pt, font=\tiny]

@@

On the other hand, we argue that a concurrency theory of rewriting can
be developed for well-switching rewriting systems, recovering, sometimes in
weakened form, results that hold in the linear case.
% Specifically, the Church-Rosser theorem is fully recovered.
%
The development has to face a main technical obstacle. In the
literature on linear rewriting systems, a central result is the
characterisation of switch equivalence of sequences of steps based on
what are called processes or consistent
permutations~\cite{CMR:GP,Handbook,BaldanCHKS06,heindel2009category}.
In turn this is the key to derive some
% basic but
fundamental
properties for linear rewriting,
% systems
namely
\begin{enumerate}
\item \emph{globality of independence}: if two independent steps are
  moved forward or backward due to the application of switchings to a
  sequence, they remain independent;
\item \emph{consistency of switching}: when transforming
  sequences of rewriting steps into equivalent ones by switching independent steps, the
  result does not depend on the order in which switchings are applied, but
  only on the associated permutation.
\end{enumerate}
Properties (1) and (2) allow one to deduce a canonical form for equivalence proofs 
between sequences of rewriting steps seen as concurrent computations.

@@

\emph{Synopsis.}
% The paper has the following structure. 
In \S\ref{sec:ade} we recall the basic 
definitions of DPO rewriting for $\mathcal{M}$-adhesive categories. 
In \S\ref{sec:equi} we present the key novelties of our proposal,
first introducing the classical notion of independence and 
a novel axiomatic notion of switchability, further showing that the
former fails to imply the latter for left-linear rewriting systems.
We then define a subclass of left-linear rewriting systems
that we call well-switching rewriting systems,
and we prove that all the classical results holding for linear rewriting systems
can be recovered for such a class.
Finally, in \S\ref{conclusions} we outline possible directions for future work.
%
The appendices
% collect additional data,
recall basic results on adhesive categories (Appendix~\ref{app:ade}),
present \emph{strong enforcing} rewriting systems, a class of systems
where the local Church-Rosser Theorem holds (Appendix~\ref{app:fill}),
and give full proofs of the results
(Appendix~\ref{omitted}).

\section{$\mathcal{M}$-adhesive categories and rewriting systems}\label{sec:ade}
This first section recalls some definitions and the basic theory of \emph{$\mathcal{M}$-adhesive categories} \cite{azzi2019essence,ehrig2012,ehrig2014adhesive,lack2005adhesive,heindel2009category}.
%
%Throughout the paper
Given a category $\X$ we will not distinguish
notationally between $\X$ and its class of objects, so that
``$X\in \X$'' means that $X$ is an object of $\X$.
We will use $\mor(\X)$, $\mon(\X)$ and $\reg(\X)$ denote the class of
all arrows, monos and regular monos of $\X$, respectively.  Given an
integer $n\in \mathbb{Z}$, $[0,n]$ denotes the set of natural numbers
less than or equal to $n$; in particular, $[0,n]=\emptyset$ whenever
$n<0$.

@@

    A $\mor(\X)$-Van Kampen square will be called simply \emph{Van
      Kampen} and similarly a $\mor(\X)$-stable square will be called
    \emph{stable}.
  \end{definition}

@@

  \begin{figure}[t]
    \input{Figure/derD.tex}
    \caption{The derivation $\der{D}$.}
    \label{fi:derD}
  \end{figure}

  \begin{figure}[b]
    \input{Figure/derE.tex}
    \caption{The derivation $\der{E}$.}
    \label{fi:derE}
  \end{figure}
\end{example}

@@

We first recall some %concepts and
notation regarding permutations. 
% \begin{definition}
A \emph{permutation} on
$\interval[0]{n}$ is a bijection
$\sigma : \interval[0]{n} \to \interval[0]{n}$. We call it a
\emph{transposition} when there are $i, j \in \interval{n}$,
$i \neq j$ such that $\sigma(i)=j$, $\sigma(j) = i$, and
$\sigma(k) = k$ otherwise. Such a transposition is
% normally
denoted as
$\transp{i}[j]$ and we use the letter $\nu$ for transpositions.
% 
Given a permutation $\perm$ on $\interval[0]{n}$, an \emph{inversion}
for $\sigma$ is a pair $(i,j)$ such that $i<j$ and
$\sigma(j)< \sigma(i)$. We write $\inv{\sigma}$ for the set of
inversions of $\perm$.
%\end{definition}

@@

Even if we work in strong enforcing rewriting systems where sequential
independence ensures switchability, when dealing with left-linear
rules there is a further -- and possibly more serious -- issue, 
namely the fact that there can be more than one independence pair between
the same derivations (cfr. \Cref{rem:uni}). This hinders the very idea of using
sequential independence as a basis of a theory of concurrency for
rewriting systems, since exchanges performed using different
independence pairs may lead to derivations that are not abstraction
equivalent, and thus one would equate computations that should
definitively be taken apart, as shown in the example below.

>>>>

\usepackage{pgf}
\usepackage{tikz}
\usepackage{tikz-cd}
\usetikzlibrary{arrows,shapes,snakes,automata,backgrounds,petri,fit,positioning,calc}
\tikzstyle{node}=[circle, draw=black, minimum size=1mm, inner sep=1.5pt, font=\tiny]

@@

On the other hand, we argue that a concurrency theory of rewriting can
be developed for well-switching rewriting systems, by recovering, 
even if sometimes in
weakened form, most of the results that hold in the linear case.
% Specifically, the Church-Rosser theorem is fully recovered.
%
The development has to face a main technical obstacle. In the
literature on linear rewriting systems, a central result is the
characterisation of switch equivalence of sequences of steps based on
what are called processes or consistent
permutations~\cite{CMR:GP,Handbook,BaldanCHKS06,heindel2009category}.
In turn this is the key to derive some
% basic but
fundamental
properties for linear rewriting,
% systems
namely
\begin{enumerate}
\item \emph{globality of independence}: if two independent steps are
  moved forward or backward due to the application of switchings to a
  sequence, they remain independent;
\item \emph{consistency of switching}: when transforming
  sequences of rewriting steps into equivalent ones by switching independent steps, the
  result does not depend on the order in which switchings are applied, but
  only on the associated permutation.
\end{enumerate}
Properties (1) and (2) allow one to deduce a canonical form for equivalence proofs 
between sequences of rewriting steps seen as concurrent computations.

@@

\emph{Synopsis.}
% The paper has the following structure. 
In \S\ref{sec:ade} we recall the basic 
definitions of DPO rewriting for $\mathcal{M}$-adhesive categories. 
In \S\ref{sec:equi} we present the key novelties of our proposal,
introducing the standard notion of independence and 
a novel axiomatic notion of switchability, showing that the
former fails to imply the latter for left-linear rewriting systems.
We then define a subclass of left-linear rewriting systems
that we call well-switching rewriting systems,
and we prove that the classical results for linear rewriting systems
are recovered for such a class.
Finally, in \S\ref{conclusions} we outline directions for future work.
%
The appendices
% collect additional data,
recall basic results on adhesive categories (Appendix~\ref{app:ade}),
present \emph{strong enforcing} rewriting systems, a class of systems
where the local Church-Rosser Theorem holds (Appendix~\ref{app:fill}),
and give proofs of the results
(Appendix~\ref{omitted}).

\section{$\mathcal{M}$-adhesive categories and rewriting systems}\label{sec:ade}
This  section recalls the basic theory of \emph{$\mathcal{M}$-adhesive categories} \cite{azzi2019essence,ehrig2012,ehrig2014adhesive,lack2005adhesive,heindel2009category}.
%
%Throughout the paper
Given a category $\X$ we will not distinguish
notationally between $\X$ and its class of objects, so
``$X\in \X$'' means that $X$ is an object of $\X$.
We let $\mor(\X)$, $\mon(\X)$ and $\reg(\X)$ denote the class of
all arrows, monos and regular monos of $\X$, respectively.  Given an
integer $n\in \mathbb{Z}$, $[0,n]$ denotes the set of natural numbers
less than or equal to $n$; in particular, $[0,n]=\emptyset$ if
$n<0$.

@@

    A $\mor(\X)$-Van Kampen square is called  \emph{Van
      Kampen} and a $\mor(\X)$-stable square  \emph{stable}.
  \end{definition}

@@

  \begin{figure}[t]
    \input{Figure/derD.tex}
    \caption{The derivation $\der{D}$.}
    \label{fi:derD}
  \end{figure}
  \begin{figure}[b]
    \input{Figure/derE.tex}
    \caption{The derivation $\der{E}$.}
    \label{fi:derE}
  \end{figure}
\end{example}

@@

We  recall some %concepts and
notation on permutations. 
% \begin{definition}
A \emph{permutation} on
$\interval[0]{n}$ is a bijection
$\sigma : \interval[0]{n} \to \interval[0]{n}$. It is a
\emph{transposition} $\nu$ if there are $i, j \in \interval{n}$,
$i \neq j$ such that $\sigma(i)=j$, $\sigma(j) = i$, and
$\sigma(k) = k$ otherwise; it is denoted as $\transp{i}[j]$.
%Such a transposition is
%% normally
%denoted as
%$\transp{i}[j]$ and we use the letter $\nu$ for transpositions.
%% 
Given a permutation $\perm$, % on $\interval[0]{n}$
an \emph{inversion}
for $\sigma$ is a pair $(i,j)$ such that $i<j$ and
$\sigma(j)< \sigma(i)$; $\inv{\sigma}$ denotes the set of
inversions of $\perm$.
%\end{definition}

@@

Even if we work in strong enforcing rewriting systems where sequential
independence ensures switchability, when dealing with left-linear
rules there is a further, possibly more serious issue, 
namely that there can be more than one independence pair between
the same derivations (cfr. \Cref{rem:uni}). This hinders the very idea of using
sequential independence as a basis of a theory of concurrency for
rewriting systems, since exchanges performed using different
independence pairs may lead to derivations that are not abstraction
equivalent, thus equating computations that should
definitively be taken apart, as shown in the example below.
